---
title: "Analyses for Manuscript 'English and Spanish letter naming in bilingual children in English-only and dual language contexts' by Sanabria et al." 
author: "Julian M. Siebert"
date: 'Last updated on: `r Sys.Date()`'
format:
  html:
    toc: true              # enable table of contents
    toc-location: left     # make it float on the right
    toc-floating: true     # floating TOC (sidebar)
    number-sections: true  # optional: auto-number sections
    code-fold: true        # optional: allow folding code blocks
    page-layout: full      # page width: 'full', 'centered', or 'custom'
    theme: default         # optional: or pick a custom theme
    css: styles.css        # optional: use custom CSS if needed
    toc-depth: 3           # optional: depth of TOC (1–6)
    code-tools: true       # optional: adds show/hide/copy buttons for code
    collapse-section: true # allow section folding
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set global chunk options
knitr::opts_chunk$set(
  echo = FALSE,       # Show the code
  warning = FALSE,    # Suppress warnings
  message = FALSE,    # Suppress messages
  cache = FALSE       # Store results
)

set.seed(1)

options(tinytex.verbose = TRUE)

library(lme4)
library(gtsummary)
library(flextable)
library(readxl)
library(ggpubr)
library(effectsize)
library(modelsummary)
library(tidyverse)

# set themes
theme_set(theme_classic())
theme_update(
  panel.background = element_rect(fill = 'transparent'),
  plot.background = element_rect(fill = 'transparent', color = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  legend.background = element_rect(fill = 'transparent', color = NA),
  legend.position = "bottom"
)

```

```{r functions}
# insert preferred plot function
plot.accuracy <- function(data) {
  plot <- data %>%
    pivot_longer(cols = c(contains("pos")),
                 names_to = "item",
                 values_to = "score") %>% 
    mutate(item = str_sub(item, 7, 9)) %>% 
    rbind(
      data %>%
        pivot_longer(cols = c(contains("pos")),
                     names_to = "item",
                     values_to = "score") %>% 
        mutate(item = str_sub(item, 7, 9)) %>% 
        mutate(lang_inst = "Overall")
    ) %>% 
    group_by(item, lang_inst) %>% 
    summarise(
      n  = sum(!is.na(score)),
      m  = round(mean(score, na.rm = TRUE), 2),
      sd = round(sd(score, na.rm = TRUE), 2),
      se = sd / sqrt(n),
      .groups = "drop"
    ) %>% 
    unique() %>% 
    left_join(
      data %>% 
        pivot_longer(cols = c(contains("pos")),
                     names_to = "item",
                     values_to = "score") %>% 
        mutate(item = str_sub(item, 7, 9)) %>% 
        group_by(item) %>% 
        summarise(
          m_all  = round(mean(score, na.rm = TRUE), 2),
          .groups = "drop"
        ) %>% 
    unique() 
    ) %>% 
    ggplot(aes(x = reorder(item, -m_all), y = m, colour = lang_inst)) +
    # geom_errorbar(aes(ymin = m - se, ymax = m + se), width = 0.2) +
    geom_line(aes(group = item), colour = "gray50") +
    geom_point(aes(y = m_all), size = 1.5, colour = "gray30") +
    geom_point(size = 1.5) +
    # geom_line(aes(y = m_all)) +
    labs(x = " ",
         y = "Mean Accuracy (Proportion Correct)",
         colour = "Instructional Context"
    ) +
    scale_colour_manual(values = c("English" = "#0072B2", "Bilingual" = "#D55E00", "Overall" = "gray30")) +
    guides(colour = guide_legend(override.aes = list(size = 4))) +
    theme(
      panel.grid.major.x = element_line(color = "gray90"),
      axis.title.x = element_blank()
      )
  plot
}

plot.feature <- function(data, feature, featurename) {
  
  df_name <- deparse(substitute(data))
  
  if (featurename %in% c("Class", "Case", "Letter Name Similarity Across Languages",
                         "Name Starts w/ Most Freq. Sound", "Name Has CV Structure")) {
    
    feature_vals <- as.character(pull(data, {{feature}}))
    comparisons <- combn(unique(feature_vals), 2, simplify = FALSE)
    # comparisons <- setdiff(comparisons, "Exact")
  
  plot <- data %>% 
    unique() %>% 
    ggplot(aes(x = {{feature}}, y = diff, colour = class))  +
    # geom_jitter(width = .2, alpha = .5) +
    geom_text(aes(label = item), 
              position = position_jitter(width = 0.3, height = 0), 
              alpha = 0.9, size = 3) +
    geom_boxplot(width = .1, fill = "white", colour = "black", alpha = .6) +
    # geom_signif(comparisons = comparisons,
    #             annotations = "n.s.",
    #             map_signif_level = TRUE,
    #             textsize = 3, 
    #             y_position = 0.98,
    #             colour = "black") + 
    facet_grid(cols = vars(lang.inst)) +
    labs(x = paste(featurename),
         y = "Mean Accuracy (Proportion Correct)"
         # colour = "Case"
    ) +
    theme(legend.position = "none") +
      scale_colour_manual(values = c("Consonant" = "black", "Vowel" = "darkred"))
  
  } else if (featurename %in% c("Log(Frequency)", "Visual Distractors (Average)")) {
    
    plot <- df.diff.lnf.en %>% 
      unique() %>% 
      ggplot(aes(x = {{feature}}, y = diff, colour = class))  +
      # ggplot(aes(x = freq, y = diff))  +
      geom_text(aes(label = item)) +
      facet_grid(cols = vars(lang.inst)) +
      labs(x = paste(featurename),
           y = "Mean Accuracy (Proportion Correct)",
      ) +
      theme(legend.position = "none") +
      scale_colour_manual(values = c("Consonant" = "black", "Vowel" = "darkred"))
    
  } else if (featurename == "Consonant Acquisition Order") {
    
    # if (grepl(df_name == "df.diff.lsf.en")) {
    # }
    plot <- data %>% 
      unique() %>% 
      ggplot(aes(x = {{feature}}, y = diff))  +
      geom_text(aes(label = item), 
                position = position_jitter(width = 0.2, height = 0), 
                alpha = 0.8, size = 3) +
      facet_grid(cols = vars(lang.inst)) +
      labs(x = paste(featurename),
           y = "Mean Accuracy (Proportion Correct)"
      ) +
      annotate("rect", xmin = 0.5, xmax = 1.5, ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "lightgray") +
      annotate("rect", xmin = 2.5, xmax = 3.5, ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "lightgray") +
      annotate("rect", xmin = 4.5, xmax = 5.5, ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "lightgray") +
      scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6))
  }
  plot +
    scale_y_continuous(limits = c(0, 1)) +
  theme(axis.title.y = element_blank(),
        axis.title = element_text(size = 8),
        axis.text = element_text(size = 6),
        strip.text = element_text(size = 6),    # facet labels
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        plot.title = element_text(size = 8)
  )
}

plot.all <- function(data) {
  p1 <- plot.feature(data, case, "Case")
  p2 <- plot.feature(data, class, "Class")
  p3 <- plot.feature(data, cross.lg.vis.sim, "Letter Name Similarity Across Languages")
  p4 <- plot.feature(data, cons.acq.order, "Consonant Acquisition Order")
  p5 <- plot.feature(data, log_freq, "Log(Frequency)")
  p6 <- plot.feature(data, avg.vis.sim, "Visual Distractors (Average)")
  p7 <- plot.feature(data, name.start.common, "Name Starts w/ Most Freq. Sound")
  p8 <- plot.feature(data, name.cv, "Name Has CV Structure")
  print(
  ggarrange(
    p1, p2, p3, p4, p5, p6, p7, p8,
    labels = "AUTO",
  nrow = 2, ncol = 4)
)
}

```

```{r load-data}
df.codes.lnf.en <- read_csv("data/df-codes-lnf-en.csv")
df.codes.lnf.es <- read_csv("data/df-codes-lnf-es.csv")

# LNF_E ======
df.lnf.en <- read_csv("data/df-lnf-en.csv")
df.diff.lnf.en <- read_csv("data/df-diff-lnf-en.csv")

df.lnf.es <- read_csv("data/df-lnf-es.csv")
df.diff.lnf.es <- read_csv("data/df-diff-lnf-es.csv")

df.sample <- read_csv("data/df-sample.csv")
```

# Sample

```{r}
#| label: tbl-sample
#| tbl-cap: "Sample Characteristics by Instructional Context and Task \\newline"

df.sample %>% 
  filter(student_id %in% c(unique(df.lnf.en$student_id), unique(df.lnf.es$student_id))) %>% 
  select(-c(student_id, event, testdate, format, language, district, school, num_correct, num_named, grade)) %>% 
  mutate(lang_inst = paste0(lang_inst, " Instruction"),
         lang_inst = factor(lang_inst, levels = c("English Instruction", "Bilingual Instruction"))) %>% 
  tbl_strata(strata = "lang_inst",
             .tbl_fun =
               ~ .x |>
               tbl_summary(by = "task",
                           label = list(ethnicity = "Hispanic/Latin(o/a)",
                                        race = "Race",
                                        ever_disability = "Ever IEP/504 Plan",
                                        el = "English Proficiency Designation",
                                        gender = "Gender",
                                        home_lang = "Home Language"
                                        )
                           )
             )
```

\FloatBarrier

# English Letter Naming Fluency

## Item Difficulty (Proportion Correct) By Letter & Instructional Context

```{r out.width='100%', fig.height=4}
#| label: fig-acc-lnf-en
#| fig-cap: "Mean Accuracy (Proportion of Correct Responses) by Instructional Context for the English Letter Naming Fluency Task, Ranked By Overall Mean Accuracy."
plot.accuracy(df.lnf.en)
```

\newpage

```{=html}
<div style="text-align: right; margin-bottom: 0.5em;">
  <button class="btn btn-outline-secondary btn-sm" type="button" data-bs-toggle="collapse" data-bs-target="#lnf-en-table">
    ▼ Show/Hide Table
  </button>
</div>
<div class="collapse" id="lnf-en-table">
```

```{r tbl-acc-lnf-en}
#| label: tbl-acc-lnf-en
#| tbl-cap: "Mean Accuracies for the English Letter Naming Fluency Task."
df.lnf.en %>%
  pivot_longer(cols = c(contains("pos")),
               names_to = "item",
               values_to = "score") %>% 
  mutate(item = str_sub(item, 7, 9)) %>% 
  rbind(
    df.lnf.en %>%
      mutate(lang_inst = "All") %>% 
      pivot_longer(cols = c(contains("pos")),
                   names_to = "item",
                   values_to = "score") %>% 
      mutate(item = str_sub(item, 7, 9))
  ) %>% 
  group_by(item, lang_inst) %>% 
  summarise(n  = sum(!is.na(score)),
            m  = round(mean(score, na.rm = TRUE), 2),
            sd = round(sd(score, na.rm = TRUE), 2)
  ) %>% 
  pivot_wider(names_from = lang_inst, values_from = c(n, m, sd)) %>% 
  select(c(item, contains("All"), contains("English"), contains("Bilingual"))) %>% 
  arrange(desc(`m_All`)) %>% 
  flextable() %>%
  set_header_labels(
    item = "Item",
    n_All = "n", m_All = "M", sd_All = "SD",
    n_English = "n", m_English = "M", sd_English = "SD",
    n_Bilingual = "n", m_Bilingual = "M", sd_Bilingual = "SD"
  ) %>%
  add_header_row(values = c("", "Overall", "English Instruction", "Bilingual Instruction"),
                 colwidths = c(1, 3, 3, 3)) %>%
  align(align = "center", part = "all") %>%
  autofit()
```

```{r tbl-features-en}
df.diff.lnf.en %>% 
  mutate(lang = "English") %>% 
  select(-`factor(cross.lg.vis.sim, levels = c(\"Dissimilar\", \"Similar\"))`) %>% 
  rbind(df.diff.lnf.es %>% 
          mutate(lang = "Spanish")
          ) %>% 
  select(-c(lang.inst, diff)) %>% 
  unique() %>% 
  mutate(vowel = case_when(class == "Vowel" ~ 1,
                           class == "Consonant" ~ 0),
         uppercase = case_when(case == "Uppercase" ~ 1,
                               case == "Lowercase" ~ 0),
         cross.lg.vis.sim = case_when(cross.lg.vis.sim == "Similar" ~ 1,
                                      cross.lg.vis.sim == "Dissimilar" ~ 0),
         ) %>% 
  select(c(lang,
           freq,
           avg.vis.sim,
           cross.lg.vis.sim,
           uppercase,
           vowel,
           name.cv
  )
  ) %>% 
  tbl_strata(strata = lang,
             .tbl_fun =
               ~ .x %>%
               tbl_summary(
                 statistic = all_continuous() ~ "{mean} ({sd})"
               )
  )
```

</div>

\blandscape

## Descriptives of Individual Features

```{r fig.width=8, fig.height=5}
#| label: fig-features-lnf-en
#| fig-cap: "Mean Accuracy (Proportion of Correct Responses) by Instructional Context and Item Features for the English Letter Naming Fluency Task."
plot.all(df.diff.lnf.en)
```

\elandscape

\FloatBarrier

## Regressions

These are generalized linear mixed models (implemented with the `lmer` package in R), with random effects for item and student.

```{r regressions-lnf-en}
df.reg.lnf.en <- df.lnf.en %>%
      filter(!is.na(lang_inst)) %>%
      pivot_longer(cols = c(contains("pos")),
                   names_to = "item",
                   values_to = "score") %>%
      mutate(item = str_sub(item, 7, 9)) %>% 
  unique() %>% 
  left_join(df.codes.lnf.en) %>% 
  unique() %>% 
  select(-c(language, event)) %>%
  # scale continuous variables
  mutate(across(c(freq, avg.vis.sim), scale)) %>% 
  mutate(
         student_id = factor(student_id),
         item = factor(item),
         )
  
data <- df.reg.lnf.en %>% filter(lang_inst == "English")
glmm.1 <- glmer(score ~ gender + case + class + freq + avg.vis.sim + cross.lg.vis.sim + name.cv + (1 | student_id) + (1 | item),
              data = data,
              family = binomial(link = "logit"),
              control = glmerControl(optimizer = "bobyqa")
              )

data <- df.reg.lnf.en %>% filter(lang_inst == "Bilingual")
glmm.2 <- glmer(score ~ gender + case + class + freq + avg.vis.sim + cross.lg.vis.sim + name.cv + (1 | student_id) + (1 | item),
              data = data,
              family = binomial(link = "logit"),
              control = glmerControl(optimizer = "bobyqa")
              )

stdcoefs.1 <- standardize_parameters(glmm.1)
stdcoefs.2 <- standardize_parameters(glmm.2)
modelsummary(
  list("English Instruction" = glmm.1, "Bilingual Instruction" = glmm.2),
  statistic = "conf.int",
  coef_rename = c(
    "genderMale" = "Gender (Male)",
    "caseUppercase" = "Case (Upper)",
    "classConsonant" = "Class (Consonant)",
    "freq" = "Frequency",
    "avg.vis.sim" = "Visual Distractors (Average)",
    "cross.lg.vis.simSimilar" = "Letter Name Similarity Across Languages (Similar)",
    "cross.lg.vis.simExact" = "Letter Name Similarity Across Languages (Exact)",
    "name.start.commonNo" = "Letter Name Starts w/ Most Frequent Sound (No)",
    "name.cvNo" = "Letter Name Has CV Structure (No)"
  ),
  conf_format = "[{conf.low}, {conf.high}]",
  coef_override = list("English instruction" = stdcoefs.1, "Bilingual Instruction" = stdcoefs.2),
  stars = c('*' = 0.05, '**' = 0.01, '***' = 0.001),
  gof_omit = "AIC|BIC"
)
```


## Additional Analyses

### Checking Balance of Item Features by CV-Name (Table)

```{r cv-balancing-analyses-en}
df.temp <- df.diff.lnf.en %>% 
  select(-c(lang.inst, diff)) %>% 
  unique() %>% 
  select(c(name.cv, case, class, cross.lg.vis.sim, freq, avg.vis.sim))

df.temp %>% 
  tbl_summary(by = name.cv) %>% 
  add_p(
    test = list(
      case ~ "chisq.test",
      class ~ "chisq.test",
      cross.lg.vis.sim ~ "chisq.test",
      all_continuous() ~ "t.test"
    ),
    test.args = all_tests("t.test") ~ list(var.equal = TRUE)
  ) %>% 
  add_stat(
    fns = list(
      all_continuous() ~ function(data, variable, by, ...) {
        t.test(data[[variable]] ~ data[[by]], var.equal = TRUE)$statistic
      },
      all_categorical() ~ function(data, variable, by, ...) {
        chisq.test(table(data[[variable]], data[[by]]))$statistic
      }
    )
  ) 
```

### Direct Chi-squared Tests

```{r}
with(df.temp, chisq.test(table(case, name.cv)))
with(df.temp, chisq.test(table(class, name.cv)))
with(df.temp, chisq.test(table(cross.lg.vis.sim, name.cv)))
```

### Differences in Frequency and Visual Similarity by CV-Name

```{r t-tests-cv-en}
t.test(freq ~ name.cv, data = df.temp, var.equal = TRUE)
t.test(avg.vis.sim ~ name.cv, data = df.temp, var.equal = TRUE)
```

### Comparing Scores by Instructional Context

```{r score-comparison-by-context-en}
t.test(num_correct ~ lang_inst, data = df.lnf.en, var.equal = TRUE)
```


# Spanish Letter Naming Fluency

## Item Difficulty (Proportion Correct) By Letter & Instructional Context

```{r out.width='100%', fig.height=4}
#| label: fig-acc-lnf-es
#| fig-cap: "Mean Accuracy (Proportion of Correct Responses) by Instructional Context for the Spanish Letter Naming Fluency Task, Ranked By Overall Mean Accuracy."
plot.accuracy(df.lnf.es)
```

\newpage

```{=html}
<div style="text-align: right; margin-bottom: 0.5em;">
  <button class="btn btn-outline-secondary btn-sm" type="button" data-bs-toggle="collapse" data-bs-target="#lnf-es-table">
    ▼ Show/Hide Table
  </button>
</div>
<div class="collapse" id="lnf-es-table">
```

```{r tbl-acc-lnf-es}
#| label: tbl-acc-lnf-es
#| tbl-cap: "Mean Accuracies for the Spanish Letter Naming Fluency Task."
df.lnf.es %>%
  pivot_longer(cols = c(contains("pos")),
               names_to = "item",
               values_to = "score") %>% 
  mutate(item = str_sub(item, 7, 9)) %>% 
  rbind(
    df.lnf.es %>%
      mutate(lang_inst = "All") %>% 
      pivot_longer(cols = c(contains("pos")),
                   names_to = "item",
                   values_to = "score") %>% 
      mutate(item = str_sub(item, 7, 9))
  ) %>% 
  group_by(item, lang_inst) %>% 
  summarise(n  = sum(!is.na(score)),
            m  = round(mean(score, na.rm = TRUE), 2),
            sd = round(sd(score, na.rm = TRUE), 2)
  ) %>% 
  pivot_wider(names_from = lang_inst, values_from = c(n, m, sd)) %>% 
  select(c(item, contains("All"), contains("English"), contains("Bilingual"))) %>% 
  arrange(desc(`m_All`)) %>% 
  flextable() %>%
  set_header_labels(
    item = "Item",
    n_All = "n", m_All = "M", sd_All = "SD",
    n_English = "n", m_English = "M", sd_English = "SD",
    n_Bilingual = "n", m_Bilingual = "M", sd_Bilingual = "SD"
  ) %>%
  add_header_row(values = c("", "Overall", "English Instruction", "Bilingual Instruction"),
                 colwidths = c(1, 3, 3, 3)) %>%
  align(align = "center", part = "all") %>%
  autofit()
```

</div>

\blandscape

## Descriptives of Individual Features

```{r fig.width=8, fig.height=5}
#| label: fig-features-lnf-es
#| fig-cap: "Mean Accuracy (Proportion of Correct Responses) by Instructional Context and Item Features for the Spanish Letter Naming Fluency Task."
plot.all(df.diff.lnf.es)
```

\elandscape
\FloatBarrier

## Regressions

```{r regressions-lnf-es}
df.reg.lnf.es <- df.lnf.es %>%
      filter(!is.na(lang_inst)) %>%
      pivot_longer(cols = c(contains("pos")),
                   names_to = "item",
                   values_to = "score") %>%
      mutate(item = str_sub(item, 7, 9)) %>% 
  unique() %>% 
  left_join(df.codes.lnf.es) %>% 
  unique() %>% 
  select(-c(language, event)) %>%
  # scale continuous variables
  mutate(across(c(freq, avg.vis.sim), scale)) %>% 
  mutate(student_id = factor(student_id),
         item = factor(item),
         )
  
glmm.1 <- glmer(score ~ gender + case + class + freq + avg.vis.sim + cross.lg.vis.sim +  name.cv + (1 | student_id) + (1 | item),
              data = df.reg.lnf.es %>% filter(lang_inst == "English"),
              family = binomial(link = "logit"),
              control = glmerControl(optimizer = "bobyqa")
              )

glmm.2 <- glmer(score ~ gender + case + class + freq + avg.vis.sim + cross.lg.vis.sim +  name.cv + (1 | student_id) + (1 | item),
              data = df.reg.lnf.es %>% filter(lang_inst == "Bilingual"),
              family = binomial(link = "logit"),
              control = glmerControl(optimizer = "bobyqa")
              )

stdcoefs.1 <- standardize_parameters(glmm.1)
stdcoefs.2 <- standardize_parameters(glmm.2)
modelsummary(
  list("English instruction" = glmm.1, "Bilingual Instruction" = glmm.2),
  statistic = "conf.int",
  coef_rename = c(
    "genderMale" = "Gender (Male)",
    "caseUppercase" = "Case (Upper)",
    "classConsonant" = "Class (Consonant)",
    "freq" = "Frequency",
    "avg.vis.sim" = "Visual Distractors (Average)",
    "cross.lg.vis.simSimilar" = "Letter Name Similarity Across Languages (Similar)",
    "cross.lg.vis.simExact" = "Letter Name Similarity Across Languages (Exact)",
    "name.start.commonNo" = "Letter Name Starts w/ Most Frequent Sound (No)",
    "name.cvNo" = "Letter Name Has CV Structure (No)"
  ),
  conf_format = "[{conf.low}, {conf.high}]",
  coef_override = list("English Instruction" = stdcoefs.1, "Bilingual Instruction" = stdcoefs.2),
  stars = c('*' = 0.05, '**' = 0.01, '***' = 0.001),
  gof_omit = "AIC|BIC"
)
```
## Additional Analyses

### Checking Balance of Item Features by CV-Name (Table)

```{r cv-balancing-analyses-es}
df.temp <- df.diff.lnf.es %>% 
  select(-c(lang.inst, diff)) %>% 
  unique() %>% 
  select(c(name.cv, case, class, cross.lg.vis.sim, freq, avg.vis.sim))

df.temp %>% 
  tbl_summary(by = name.cv) %>% 
  add_p(
    test = list(
      case ~ "chisq.test",
      class ~ "chisq.test",
      cross.lg.vis.sim ~ "chisq.test",
      all_continuous() ~ "t.test"
    ),
    test.args = all_tests("t.test") ~ list(var.equal = TRUE)
  ) %>% 
  add_stat(
    fns = list(
      all_continuous() ~ function(data, variable, by, ...) {
        t.test(data[[variable]] ~ data[[by]], var.equal = TRUE)$statistic
      },
      all_categorical() ~ function(data, variable, by, ...) {
        chisq.test(table(data[[variable]], data[[by]]))$statistic
      }
    )
  ) %>% 
  modify_header(stat_1 = "**Test statistic**")
```

### Direct Chi-squared Tests

```{r}
with(df.temp, chisq.test(table(case, name.cv)))
with(df.temp, chisq.test(table(class, name.cv)))
with(df.temp, chisq.test(table(cross.lg.vis.sim, name.cv)))
```

### Differences in Frequency and Visual Similarity by CV-Name

```{r  t-tests-cv-es}
t.test(freq ~ name.cv, data = df.temp, var.equal = TRUE)
t.test(avg.vis.sim ~ name.cv, data = df.temp, var.equal = TRUE)
```

### Comparing Scores by Instructional Context

```{r score-comparison-by-context-es}
t.test(num_correct ~ lang_inst, data = df.lnf.es, var.equal = TRUE)
```
